Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-09-12 concurrency/CPP 2024/paper.tex
Shengyi's comments, more reviewer comments

\section{Introduction}
\label{sec:introduction}
Krishna et al. ~\cite{templates} proposed concurrent search structure templates as a method for proving the correctness of concurrent data structures \emph{compositionally}, separating the proof of a concurrent access method (e.g., optimistic concurrency, hand-over-hand locking, forwarding via internal links) from the proof of the underlying data structure (e.g., linked list, hashtable, B-tree). The concurrency ``templates'' are verified parametrically over data structure operations, and the data structure operations are verified without any reference to concurrency. In theory, this could allow us to prove the correctness of $n$ (single-threaded) data structure implementations and $m$ concurrency patterns, and immediately obtain $n \times m$ verified concurrent data structures. In practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model. %\wm{Maybe say more about the idea that they only built a proof of concept, and there are still important details not worked out/only tested on ad-hoc examples.}

@@

A concurrent search structure is a data structure that supports three operations: \lstinline{insert}, \lstinline{lookup}, and \lstinline{delete}. It is intended to implement a map from keys to values, and to behave correctly when accessed simultaneously by any number of threads. In separation logic, we can prove correctness of a search structure by showing that its operations satisfy the following logically atomic specifications, where $\nodeboxrep$ is a per-thread handle to the data structure and $\treerep$ is a shared assertion linking the values in memory to an abstract map $m$ from keys to values:
\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{insert(r, k, v)}\ \\
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\end{mathpar}
\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{lookup(r, k)}\ \\
	{\color{blue}
		\left\langle \texttt{v}. \ \
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m \land m(\texttt{k}) = \texttt{v}
		\right\rangle
	}
\end{mathpar}
\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{delete(r, k)}\ \\
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
		\right\rangle
	}
\end{mathpar}
In other words, each operation takes effect atomically on the abstract map $m$, reading and/or updating it as appropriate to the operation. Krishna et al. use flow interfaces~\cite{krishna2017flow} to implement the $\treerep$ assertion connecting the data structure implementation to the abstract map, but this is not fundamental to the approach: any definition of $\nodeboxrep$ and $\treerep$ that can be used to prove the specifications (and is nontrivial, i.e., holds on the initial state of the data structure) will yield a correct data structure.

@@

\begin{mathpar}
	{\color{blue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ m \right\rangle}\ \\ \texttt{acquire}\left(n\texttt{->lock}\right)\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m\right\rangle}\\
	{\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m \right\rangle}\ \\ \texttt{release}(n\texttt{->lock})\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ m\right\rangle}
\end{mathpar}

@@

The resources $\mathsf{R}$ contained in a node depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. Formally, the lock invariant is defined by 
\begin{align*}&\mathsf{node\_contents}(n, c, \mathsf{range}) \triangleq \\ 
	&\qquad\mathsf{ghost\_node}(n, c,\mathsf{range}) \ \ast \ \mathsf{node\_data}(n, c) \\
	&\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})\end{align*}
where the definition of $\mathsf{node\_data}$ is supplied by the target data structure. The contents $c$ of a node can be either a key-value pair $(k, v)$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys).

@@

The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
\begin{mathpar}
	{\color{blue}
		\fforall \  m. \left\langle
		\begin{array}{c}
			\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ m
		\end{array}
		\right\rangle
	}
	\vspace{-6pt}\\ \vspace{-6pt}
	\texttt{traverse(pn, k)} 
	\\
	{\color{blue}
		\left\langle \mathit{res}. \ \exists \  n', v.
		\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \\ \texttt{k} \in \mathsf{range}\ \ast\ 
			\\ 
			(\mathsf{if} \ \mathit{res} \\ 
			\ \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
			\\ 
			\ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
		\end{array}
		\Bigg| \ \treerep\ m \
		\right\rangle
	}
\end{mathpar}
As before, the \texttt{traverse} function will navigate the structure and return a node $n'$ whose range includes \texttt{k}, as well as acquiring $n'$'s lock and returning its contents. Also as before, the Boolean $\mathit{res}$ indicates whether $n'$ is empty or contains the key \lstinline{k}.

@@

The total lines of code required were about the same as for the original VST verification of the BST. The vast majority of the proof effort was in the templates, in both \lstinline{traverse} and the top-level data structure operations. This is fairly encouraging, since this proof effort should be reusable for other data structures that use the same templates. For lock-coupling deletion, we retained the original (non-templated) proofs; we did not verify give-up deletion, since optimistic deletion is considerably harder to reason about and did not stand to benefit from the template approach.
%Even though the proof for the \texttt{insert} function within the lock-coupling templates was line-of-code (LoC) intensive, future enhancements could significantly streamline this process. Otherwise, the construction of proofs for the \texttt{traverse} function, in the contexts of both lock-coupling and give-up templates, necessitated a considerable amount of time and effort.
%However, we expect that functions such as \texttt{traverse}, \texttt{insert}, and \texttt{lookup} will be reusable with only minor adjustments needed.  The methods \texttt{findNext} and \texttt{insertOp} may also be adapted to other data structures, such as linked lists, with some additional effort. 
Proving the correctness of e.g. a lock-coupling linked list should be as simple as defining a new \lstinline{node} struct, and then implementing and verifying \lstinline{findNext} and \lstinline{insertOp} (plus the appropriate functions for deletion).
%Specifically, when attempting to prove concurrent linked lists, we plan to tweak the specifications of \texttt{findNext}, \texttt{inRange}, and \texttt{insertOp} to restructure the \texttt{node} struct. While the size of our template proofs currently approximates that of the no-template versions, we anticipate that strategic application of simple tactics for quantifier expansion could significantly streamline these proofs in the future.
The full development of our mechanization effort is available online at [link redacted for anonymity].
%\href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}.

>>>>

\section{Introduction}
\label{sec:introduction}
Krishna et al. ~\cite{templates} proposed concurrent search structure templates as a method for proving the correctness of concurrent data structures \emph{compositionally}, separating the proof of a concurrent access method (e.g., optimistic concurrency, hand-over-hand locking, forwarding via internal links) from the proof of the underlying data structure (e.g., linked list, hashtable, B-tree). The concurrency ``templates'' are verified parametrically over data structure operations, and the data structure operations are verified without any reference to concurrency. In theory, this could allow us to prove the correctness of $n$ concurrency patterns and $m$ (single-threaded) data structure implementations, and immediately obtain $n \times m$ verified concurrent data structures. In practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model. %\wm{Maybe say more about the idea that they only built a proof of concept, and there are still important details not worked out/only tested on ad-hoc examples.}

@@

A concurrent search structure is a data structure that supports three operations: \lstinline{insert}, \lstinline{lookup}, and \lstinline{delete}. It is intended to implement a map from keys to values, and to behave correctly when accessed simultaneously by any number of threads. In separation logic, we can prove correctness of a search structure by showing that its operations satisfy the following logically atomic specifications, where $\nodeboxrep$ is a per-thread handle to the data structure and $\treerep$ is a shared assertion linking the values in memory to an abstract map $m$ from keys to values:
\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{insert(r, k, v)}\ \\
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}
\end{mathpar}
\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{lookup(r, k)}\ \\
	{\color{blue}
		\left\langle \texttt{v}. \ \
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m \land m(\texttt{k}) = \texttt{v}
		\right\rangle
	}
\end{mathpar}

\begin{mathpar}
	{\color{blue}
		\fforall m.\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ m
		\right\rangle
	}
	\ \texttt{delete(r, k)}\ \\
	{\color{blue}
		\left\langle 
		\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \_ ])
		\right\rangle
	}
\end{mathpar}
In other words, each operation takes effect atomically on the abstract map $m$, reading and/or updating it as appropriate to the operation. Krishna et al. use flow interfaces~\cite{krishna2017flow} to implement the $\treerep$ assertion connecting the data structure implementation to the abstract map, but this is not fundamental to the approach: any definition of $\nodeboxrep$ and $\treerep$ that can be used to prove the specifications (and is nontrivial, i.e., holds on the initial state of the data structure) will yield a correct data structure.

@@

\begin{mathpar}
	{\color{blue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ m \right\rangle}\ \\ \texttt{acquire}\left(n\texttt{->lock}\right)\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m\right\rangle}
	\end{mathpar}
	\begin{mathpar}
	{\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m \right\rangle}\ \\ \texttt{release}(n\texttt{->lock})\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ m\right\rangle}
\end{mathpar}

@@

The resources $\mathsf{R}$ contained in a node depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. Formally, the lock invariant is defined by 
\begin{align*}&\mathsf{node\_contents}(n, c, \mathsf{range}) \triangleq \\ 
	&\qquad\mathsf{ghost\_node}(n, c,\mathsf{range}) \ \ast \ \mathsf{node\_data}(n, c) \\
	&\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})\end{align*}
where the definition of $\mathsf{node\_data}$ is supplied by the target data structure. The contents $c$ of a node can be either a key-value pair $(k, v)$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys). Then $\treerep\ m$ is defined as a collection of $\mathsf{ghost\_node}$s that form a tree containing all the key-value pairs in $m$.

@@

The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
\begin{mathpar}
	{\color{blue}
		\fforall \  m. \left\langle
		\begin{array}{c}
			\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ m
		\end{array}
		\right\rangle
	}
	\vspace{-6pt}\\ \vspace{-6pt}
	\texttt{traverse(pn, k)} 
	\\
	{\color{blue}
		\left\langle \mathit{res}. \ \exists \  n', v.
		\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \\ \texttt{k} \in \mathsf{range}\ \ast\ 
			\\ 
			(\mathsf{if} \ \mathit{res} \\ 
			\ \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
			\\ 
			\ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
		\end{array}
		\Bigg| \ \treerep\ m \
		\right\rangle
	}
\end{mathpar}
As before, the \texttt{traverse} function will navigate the structure and return a node $n'$ whose range includes \texttt{k}, as well as acquiring $n'$'s lock and returning its contents. Also as before, the Boolean $\mathit{res}$ indicates whether $n'$ is empty or contains the key \lstinline{k}. Our definition of $\treerep$ for the give-up template closely follows that of Krishna et al.

@@

The total lines of code required were about the same as for the original VST verification of the BST. The vast majority of the proof effort was in the templates, in both \lstinline{traverse} and the top-level data structure operations. This is fairly encouraging, since this proof effort should be reusable for other data structures that use the same templates. For lock-coupling deletion, we retained the original (non-templated) proofs; we did not verify give-up deletion, since optimistic deletion is considerably harder to reason about and did not stand to benefit from the template approach.
%Even though the proof for the \texttt{insert} function within the lock-coupling templates was line-of-code (LoC) intensive, future enhancements could significantly streamline this process. Otherwise, the construction of proofs for the \texttt{traverse} function, in the contexts of both lock-coupling and give-up templates, necessitated a considerable amount of time and effort.
%However, we expect that functions such as \texttt{traverse}, \texttt{insert}, and \texttt{lookup} will be reusable with only minor adjustments needed.  The methods \texttt{findNext} and \texttt{insertOp} may also be adapted to other data structures, such as linked lists, with some additional effort. 
Proving the correctness of e.g. a lock-coupling linked list should be as simple as defining a new \lstinline{node} struct, and then implementing and verifying \lstinline{findNext} and \lstinline{insertOp} (plus the appropriate functions for deletion).
%Specifically, when attempting to prove concurrent linked lists, we plan to tweak the specifications of \texttt{findNext}, \texttt{inRange}, and \texttt{insertOp} to restructure the \texttt{node} struct. While the size of our template proofs currently approximates that of the no-template versions, we anticipate that strategic application of simple tactics for quantifier expansion could significantly streamline these proofs in the future.

Compared to prior template proofs in Iris and GRASShopper, our proof development is significantly larger (although direct comparison is difficult, since GRASShopper is an automatic prover). The biggest structural difference is that we need to directly state loop invariants for our imperative code, while the Iris proofs handle recursion with a simpler Löb-induction approach. Coming up with the loop invariants was the hardest part of the proofs. Our proofs also deal with the details of real C code: proving absence of integer overflows, accurately representing struct-and-pointer-based data structures instead of functional-programming-style structured data, etc. In exchange for this extra effort, the code we verify constitutes complete C programs that can be compiled and executed.
The full development of our mechanization effort is available online at [link redacted for anonymity].
%\href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}.
